%% acknowledgments.tex
%% Acknowledgments - say thanks to your family, co-workers etc.
%% Use inside the document
%% @author Patrick Holthaus (pholthau)


\chapter*{Acknowledgments}%
\pdfbookmark[1]{Acknowledgments}{Acknowledgments}
%\addcontentsline{toc}{chapter}{Acknowledgments}%
\thispagestyle{empty}

% intro
I never would have been able to submit this thesis without the funding, chances, help, and advice from others.
% funding bodies
First of all, I owe thanks to Britta Wrede, Franz Kummert, and Gerhard Sagerer for giving me the opportunity and funding to work in the Applied Informatics Group and the Cluster of Excellence Cognitive Interaction Technology ``CITEC'' (EXC 277) at Bielefeld University, funded by the German Research Foundation (DFG).
% supervisors
Thanks to my supervisor Franz Kummert, in particular, for guiding me during my research and taking the time for regular and fruitful discussions.
% commitee
Moreover, I want to thank Prof. Lohan for accepting the invitation to review this thesis, and Prof. Rückert and Dr. Ell for taking part in my examination board.
% colleagues
As this research is not build up from nothing, I owe thanks to the research community and open source community in general for laying the foundations.
The members of the AI Group and CITEC deserve thanks for helpful feedback, discussions and support.
In particular, I my colleagues in the CSRA project who worked with me overtime and into weekends to create and maintain the apartment and conduct studies in it.
% research participants
Furthermore, the people who agreed to participate in these studies deserve special thanks too.
% private
I want to thank my family and friends for the support they give me.
For being there for me and for making my life worth living.
% birte
However, the biggest thanks of all deserves Birte.
My schoolmate, fellow student, colleague in the CSRA and AI, reviewer, proofreader, family, friend, and partner.
She superbly juggles all of these roles and more and deserves thanks for each of them.

\cleardoublepage
